Higher-Order unification for free!

### 2024-04-16 paper/main.tex
add todo
rework 5.18

To reason about unification, we provide a description of the
\Fo{} and \Ho languages where unification variables
are first-class terms, i.e. they have a concrete syntax as
shown in \cref{code:common-terms}.
Unification variables
in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
the arguments of a higher-order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
In \Ho, the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher-order and come equipped with an
explicit scope.

% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

\begin{proof}[Proof sketch]
  We want to prove that for any $i$, $\fstep_i$ succeeds iff $\hstep_i$
  succeeds. An $\hstep_i$ is made by a unification
  $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  The interesting case is when the infolved terms are in \maybeeta.
  wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
  we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
  In all cases $X \Ue t2$ succeeds since X is a variable and
  hence assigns t2 to X.
  Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
  additional care of eta expanding t2 if needed (5.9.1).

  Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
  $s1$ with $s2$. The interesting case is when fostep succeds by
  finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
  rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
  and that is in exactly the cases in which T (from page 3) is
  not a lam (see 5.9.1) compensating for the missing rules in \Ue.

  2) If $\fstep_i$ fails to unify $s1 \Uo s2$
  it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
  and the one flam-flam we find different terms.
  (5.9.1) only adds lambdas that are forcibly consumed by
  lam-lam rule.
  \end{proof}

Note that \progressetadedup always unifies terms starting with exaclty one
lambda, hence lam-lma, so we can't fail due to that lambda or succed thanks
to that lambda. Hence it does not impact fidelity.

>>>>

To reason about unification, we provide a description of the
\Fo{} and \Ho languages where unification variables
are first-class terms, i.e. they have a concrete syntax as
shown in \cref{code:common-terms}.
Unification variables
in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
the arguments of a higher-order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
In \Ho, the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher-order and come equipped with an
explicit scope.
%
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

\begin{proof}[Proof sketch]\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  succeeds. A $\hstep_i$ is made by a unification $u$ and a link
  progression $p$.
  Without loss of generality,\todo{explain why} we consider a \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$, which is compiled into the \linketa $\linketaM{}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to treat the progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$, in this case, the unification problem becomes
          $X \Ue t_2$, which assigns $X$ to $t_2$.
          After the unification, \progressetaleft triggers the \linketa,
          and $\lambda x.t_1$ is unified with $t_2$ (or its $\eta$-expansion when needed).
          This unification succeeds iff $s_1$ and $s_2$ unify. \progressetaleft
          aims essentially to transpose the two rules \ref{clause:eta1} and \ref{clause:eta2}
          of the object language in the meta one.
    \item $s_2 \in \maybeeta$, in this case, the unification problem becomes $X \Ue Y$ 
          with the extra link $\linketaM{}{Y}{\lambda x.t_2}$. After the unification of 
          $X$ and $Y$, \progressetadedup is fired and $t_1$ and $t_2$ are unified.
          This unification succeeds iff $s_1 \Uo s_2$ succeeds.
  \end{itemize}
  In this setting, \progressetaright does not play a critical role since \mapStore
  is a bijection; it's role is to perform variable refinement, i.e. it never fails.
  % wlog we assume s1 is in \maybeeta, s2 is not (if both are thanks to dedup we are back in this situation).
  % we assume s1 is compiled to X and $\linketaM{\Gamma}{X}{t1}$.
  % In all cases $X \Ue t2$ succeeds since X is a variable and
  % hence assigns t2 to X.
  % Rule \progressetaleft unify with \Ue X (hence t2) with t1, with the only
  % additional care of eta expanding t2 if needed (5.9.1).
% 
  % Two cases should be analyzed: 1) $\fstep_i$ succeeds to unify 
  % $s1$ with $s2$. The interesting case is when fostep succeds by
  % finding a $\rho$ such that $\rho s1 \Eo \rho s2$ requires
  % rules~\ref{clause:eta1} and \ref{clause:eta2} at the top level,
  % and that is in exactly the cases in which T (from page 3) is
  % not a lam (see 5.9.1) compensating for the missing rules in \Ue.
  % 
  % 2) If $\fstep_i$ fails to unify $s1 \Uo s2$
  % it means that, after applying rules~\ref{clause:eta1} and \ref{clause:eta2}
  % and the one flam-flam we find different terms.
  % (5.9.1) only adds lambdas that are forcibly consumed by
  % lam-lam rule.
\end{proof}

% Note that \progressetadedup always unifies terms starting with exaclty one
% lambda, hence lam-lma, so we can't fail due to that lambda or succed thanks
% to that lambda. Hence it does not impact fidelity.
